---
title: "Missing Data Approaches"
author: "cflagg"
date: "June 19, 2017"
output: html_document
---

1. Strategize
    * Two examples to keep in mind:
      * Predictors of high ozone in New York City (e.g. "air quality" data)
      * Effectiveness of drug for treating depression ("clinical" data)
    * Project Goals
        * Predictive vs. Explanatory
            * Predictive: usually creating a statistical model of some kind to predict response (Y)
                * If only interested in predictive outcomes, 
            * Explanatory: using statistical model to understand causal/mechanistic patterns
                * We want to be careful about describing patterns of cause and effect in our conclusions
    * Project Question, what's the problem?
    * Study Type
        * Observational (are there natural patterns that cause missing data?)
        * Experimental (does experimental manipulation lead to missing data?)
        * Participatory (i.e. studies about people e.g. do certain types of participants answer/not answer in certain ways?)
        
2. Understanding 'Missingness'
    * Data Structure
        * Temporal?
        * Spatial?
        * Group?
        * Response?
        * Predictor?  
    * Flavors 
        * Censoring (left, right, full)
        * Missing Completely at Random (MCAR)
        * Missing at Random (MAR)
        * Not Missing at Random (NMAR)
    * Sources   
        * Surveys: respondents not answering questions
        * Experiment: Loss due to destruction
        * Long term study: drop-out (observed start, no observed end, respondent disappears), right censored (observed beginning, no observed ending), left censored (did not observe beginning, observed ending)
        * Biased Missingness Example: sampling for species presence/absence; NA in Y always occurs in low areas that flood during the spring

3. Profiling Missingness in our Data
4. Non-Statistical Approaches
5. Statistical Approaches



```{r}
library(dplyr)
library(ggplot2)
```

## Profile missingness

* Identifying Missingness Patterns
    * Response?
    * Predictor?    
    * Temporal?
    * Spatial?
    * Group?
    
* Identify patterns of missingness to understand: 
        * Measurement Bias
            * Measurement Bias will skew model results
        * Unmeasured 'responses'
        
### Count Total Number of Missing Data Points

```{r}
# assign to an object
na_count <- apply(X = airquality, 2, is.na)

# view the first few rows of output
na_count %>% head(5)

# Count the total number of occurences where is.na==TRUE
length(na_count[na_count==TRUE])
```

### Determine Missingness per Column: Response vs. Predictors

```{r}
apply(na_count, 2, function(x) length(x[x==TRUE]))
```

* Quite a few missing response (Y) values
* A few missing predictor (X) values in Solar.R

### Is there a pattern to missing values?

* Let's visualize predictor distributions in records WITH ozone measurements
* First, subset data into two different data frames for plotting: one with zero missing Y measurements, the other only containing missing Y

```{r}
aq_data <- airquality
aq_data$time <- as.numeric(paste0(aq_data$Month + aq_data$Day)) # for quick plotting by time
present_data <- filter(aq_data, is.na(Ozone)==FALSE)
missing_data <- filter(aq_data, is.na(Ozone)==TRUE)
```

```{r}
## Records with Ozone measurements
car::scatterplotMatrix( ~ Solar.R + Wind + Temp + Month + Day, data = present_data)
```

* Now visualize predictor distributions in records WITHOUT ozone measurements, are there any visual differences between these two sub-sets?

```{r}
# records missing Ozone measurements
car::scatterplotMatrix( ~ Solar.R + Wind + Temp + Month + Day, data = missing_data)
```

* This second subset appears to have quite a few missing values in Month 6
* What would be a more alarming visual cue for measurement bias?
* What if a predictor variable is highly correlated with missingness instead?

```{r}
## Subset to missing ozone measurements
fishy <- filter(airquality, is.na(Ozone)==TRUE)
## create random wind speeds twice the mean
fishy$Wind <- rpois(n = nrow(fishy), lambda = mean(airquality$Wind)*2)

ggplot(present_data, aes(Wind)) + geom_histogram() + scale_x_continuous(limits = c(0, 25)) + ggtitle("Wind Speed with Ozone Measurements")
ggplot(fishy, aes(Wind)) + geom_histogram() + scale_x_continuous(limits = c(0, 25)) + ggtitle("Wind Speeds Missing Ozone Measurements")
```

* Here we see there is a clear pattern of missingness

### Temporal Profiling

* Are there patterns of missingness over the measurement time period?

```{r}
ggplot(present_data, aes(x = time, y = Wind)) + geom_point() + geom_line() + scale_y_continuous(limits=c(0, 20)) + scale_x_continuous(limits=c(0, 50))
ggplot(missing_data, aes(x = time, y = Wind)) + geom_point() + geom_line() + scale_y_continuous(limits=c(0, 20)) + scale_x_continuous(limits=c(0, 50))
```

### Spatial Profiling

* This data set only comes from a single monitoring station in New York City
* If we had data for multiple monitoring stations, we would want to understand

## 4. Non-Statistical Approaches
    * Record Deletion (remove records before modeling)
    * Category Merging (do multiple categories represent similar information)
        * Gender ("combine" male and female responses by ignoring gender, make inferences about treatment irrespective of male/female differences)
        * Income  
        * Experimental Treatment Groups (e.g. 1x Nitrogen addition, 2x Nitrogen addition, 3x Nitrogen addition, combine the last two levels if many data points missing)
    * Last Observation Carried Forward (LOCF): take the last available value, apply it to the next missing observation
        * Problematic if patterns are complex

## 5. Statistical Approaches
    * Simple Imputation
        * Mean/Median (replace missing value with response/predictor by group or time interval i.e. the 'group mean')
    * Multiple Imputation
        * Assume distribution model
        * Impute values per response or predictor variable
        * Repeat for each variable
        * Bootstrap imputation N times
        * Average model outputs to understand multi-variate bias
        
        
## 6. Sources

### Scholarly Papers

https://www.ncbi.nlm.nih.gov/pubmed/18823677

https://link.springer.com/article/10.1007%2Fs00265-010-1044-7

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3767219/

https://gking.harvard.edu/category/research-interests/methods/missing-data

### Technical Guides

http://www.stat.columbia.edu/~gelman/arm/missing.pdf
http://www.statmethods.net/input/missingdata.html
https://stackoverflow.com/questions/8317231/elegant-way-to-report-missing-values-in-a-data-frame
R Packages for Imputation: https://www.analyticsvidhya.com/blog/2016/03/tutorial-powerful-packages-imputing-missing-values/

### Real World Approaches (blogs)

http://www.jordicasanellas.com/data-science-blog/missing-data-impute-or-do-not-impute-r-examples
http://www.kdnuggets.com/2016/03/doing-data-science-kaggle-walkthrough-cleaning-data.html
http://www.datasciencecentral.com/profiles/blogs/how-to-treat-missing-values-in-your-data-1
https://www.r-bloggers.com/missing-values-data-science-and-r/