---
title: "phe_plots"
author: "Cody Flagg"
date: "November 19, 2015"
output: 
  html_document: 
    toc: yes
---

# A brief summary of phenology plot ideas 

## Import Data & Set Libraries
```{r warning=FALSE, message=FALSE}
library(ggplot2)
library(plyr) # always load plyr before dplyr, Mr. Wickham uses overlapping function names for some reason
library(dplyr)
library(wesanderson)

# sources: 
# get dplyr props right: http://stackoverflow.com/questions/18057081/ddply-summarise-proportional-count
# ggplot position arg: http://sape.inf.usi.ch/quick-reference/ggplot2/position
# stacked bar: http://stackoverflow.com/questions/3619067/stacked-bar-chart-in-r-ggplot2-with-y-axis-and-bars-as-percentage-of-counts
# stacked area: http://stackoverflow.com/questions/4651428/making-a-stacked-area-plot-using-ggplot2
# stacked area: http://stackoverflow.com/questions/12323060/geom-area-plot-with-areas-and-outlines-ggplot

# read in data
rm(list=ls())
options(stringsAsFactors = FALSE) 

if (file.exists('C:/Users/cflagg')) {
  pheData <- read.csv("C:/Users/cflagg/Documents/GitHub/codingSupportGroup/dataSkillsOSIS/data/NEON/phe_2014/NEON.D10.STER.DP1.10055.001.phe_statusintensity.csv", header = T)
}

if (file.exists('C:/Users/nrobinson')) {
  pheData <- read.csv("C:/Users/nrobinson/Desktop/MyDocuments/NEON_Git/codingSupportGroup/dataSkillsOSIS/data/NEON/phe_2014/NEON.D10.STER.DP1.10055.001.phe_statusintensity.csv", header = T)
}

# scramble phenophaseStatus for now, since it's all homogeneous
# DON'T RUN ON FULL DATASET
pheData$phenophaseStatus <- sample(pheData$phenophaseStatus)
```


## Calculation Check 

* Make sure the following dplyr() transforms are correct

```{r}
# these are the correct totals, but not in an ideal format
xtabs(~dayOfYear+phenophaseName+phenophaseStatus, pheData)
```

## Figures

* Looking at phenophaseName == "Leaves" only

### Figure 1 (Caren's barplot)

* x-axis = sites X dayOfYear
* y-axis = proportion of same taxonID/individualID within that phenophase (number in phenophase/total in taxonID)
* groups = phenophaseStatus (% of group in phenophase X)

```{r}
# dplyr method
# how do we pass the total individualID count without hard coding it? Coun't it before hand with a ddply
# don't want to hardcode because # of individuals could be different because of missing tags or field error
# obtain count of individualID per growthForm per dayOfYear -- this = totalID_n
pheData2 <- ddply(pheData, .(dayOfYear, growthForm), mutate, totalID_n = length(unique(individualID))) # first calculate total individuals to get
pheData2 %>% 
  dplyr::group_by(dayOfYear, phenophaseName, phenophaseStatus, growthForm,totalID_n) %>% 
  dplyr::summarize(n = length(phenophaseStatus)) %>% 
  dplyr::mutate(prop = (n/totalID_n)*100) %>% 
  # limit to beginning and end days -- full plot looks weird
  dplyr::filter(phenophaseName == "Leaves", dayOfYear %in% c(90,191)) %>% 
  ggplot(aes(x = dayOfYear, y = prop)) + geom_bar(aes(fill=phenophaseStatus), stat="identity", position = "dodge") + scale_fill_manual(values = wes_palette("Darjeeling"))
```


### Figure 2 (Lee's box & whisker)

* x-axis = phenophase (1,2,3) in single taxonID
* y-axis = days in phenophase (number of individualID/total individualID)
* group  = siteID

```{r}
# this is why the boxes are so even
xtabs(~phenophaseStatus+dayOfYear+phenophaseName, pheData2)
```

```{r}
pheData2 %>% 
  dplyr::group_by(dayOfYear, phenophaseName, phenophaseStatus, growthForm,totalID_n) %>% 
  dplyr::filter() %>%
  ggplot(aes(x = phenophaseStatus, y = dayOfYear, colour = phenophaseName, fill = phenophaseName)) + geom_boxplot()
```

### Figure 3 (Cody's line-version)
### Line Version

* x-axis = dayOfYear (full-series)
* y-axis = proportion of individuals in phenophaseStatus
* lines = phenophaseStatus (with area fill)
* facet_wrap phenophaseName?

```{r}
pheData2 %>% 
  group_by(dayOfYear, phenophaseName, phenophaseStatus, growthForm,totalID_n) %>% 
  summarize(n = length(phenophaseStatus)) %>% 
  mutate(prop = (n/totalID_n)*100) %>% 
  filter(phenophaseName == "Leaves") %>% 
  ggplot(aes(x = dayOfYear, y = prop, group = phenophaseStatus))  + geom_line(aes(colour=phenophaseStatus), size = 2)+ scale_colour_manual(values = wes_palette("Darjeeling")) 
# + facet_wrap(~phenophaseName)
```

* Would look cool if the factors were re-ordered so alpha made the graph look better

```{r}
pheData2 <- ddply(pheData, .(dayOfYear, growthForm), mutate, totalID_n = length(unique(individualID))) # first calculate total individuals to get denominator of proportion -- by day and growthForm (eventually)
pheData2 %>% 
  group_by(dayOfYear, phenophaseName, phenophaseStatus, growthForm,totalID_n) %>% 
  summarize(n = length(phenophaseStatus)) %>% 
  mutate(prop = (n/totalID_n)*100) %>% 
  filter(phenophaseName == "Leaves") %>% 
ggplot(aes(x = dayOfYear, y = prop)) + 
  geom_area(aes(group = phenophaseStatus, fill = phenophaseStatus), position = "dodge", alpha=0.5) + scale_fill_manual(values = wes_palette("Darjeeling"))

```

### Figure 4 (Natalie's impression of what Lee had in mind)

* x-axis = site
* y-axis = DOY of occurrence of single phenophase 
* group  = siteID

* NOTE - the figure below illustrates one site, there would be another next to it in the final

```{r}
#Not done using dplyr cause I'm not that familiar and wanted to churn this out real quick like...
pheData3 <- filter(pheData2, phenophaseStatus == 'yes', phenophaseName == 'Open flowers')

summary(pheData3$dayOfYear)

boxplot(summary(pheData3$dayOfYear), main="Timing of open flower phenophase per site", 
  	xlab="siteID", ylab="Day of Year of open flowers found",names=c('STER') )
mtext(side=1, 'STER')

```
