---
title: "dplyr_session2"
author: "Cody Flagg"
date: "Monday, May 04, 2015"
output: pdf_document
---

```{r echo=FALSE,warning=FALSE,message=FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
```

# Outline

1. rename() : rename columns in a data frame
2. arrange() : order rows
2. group_by() : summary statistics by groups you define, like ddply(data, grp, transform)
3. Piping/Chaining %>% : connect manipulations into a single call
  - http://www.sharpsightlabs.com/dplyr-intro-data-manipulation-with-r/
  - https://rpubs.com/justmarkham/dplyr-tutorial
  - https://rpubs.com/mkapur/ModelingProject

## Future Topics
4. SQL functions
    - Connecting to databases (http://cran.r-project.org/web/packages/dplyr/vignettes/databases.html)
    https://github.com/datacarpentry/R-dplyr-ecology/blob/gh-pages/03-data-analysis.Rmd
    https://github.com/justmarkham/dplyr-tutorial/blob/master/dplyr-tutorial.Rmd
    - Manipulating data from DBs 
5. Two-table joins : multi-table data manipulation (http://cran.r-project.org/web/packages/dplyr/vignettes/two-table.html)
6. Data Cleaning (http://stackoverflow.com/questions/23642811/replace-parts-of-a-variable-using-numeric-indices-in-dplyr-do-i-need-to-create?rq=1)

## 1. rename() - rename columns

```{r}
# Let's play with the iris dataset for a bit, RA Fisher first analyzed these data in 1936
# first print the names of the columns

data(iris)
iris

# list the column names first, so you know what you're working with
names(iris)
# The periods in each column name are kind of annoying, let's fix that
rename(iris, petLength = Petal.Length, 
  petWidth = Petal.Width, 
  sepWidth = Sepal.Width, 
  sepLength = Sepal.Length) -> iris2 # turn the assignment around, remember this for piping later

# rename and select have special arguments for finding columns
select(iris, matches("sepal", ignore.case = FALSE)) %>% head # case sensitive
select(iris, matches("sepal")) %>% head # not case sensitive

# another example - notice that it is not case sensitive
select(iris, contains("l.l")) %>% head 

# similar to a regex function i.e. stringr::sub_str()
select(iris, ends_with("th")) %>% head # the regex starts at the end of a string; if end of string doesn't explicitly begin with "th" they are skipped 

# what if you have lots of columns, named/numbered something?
# create a silly example
df1 = data.frame(x05 = rnorm(5), x06= runif(5), x07 = rpois(5, 2), type = sample(letters, 5))
# now select a few columns
select(df1, num_range("x", 5:6, width=2)) # the width argument specifies that the string is at least 2 characters wide (i.e. "01" vs. "1")
```


## 2. arrange() - re-order rows by column

```{r}
# arrange(dataframe, column1, column2, ...)
head(arrange(iris2, desc(Species), sepLength))

# the piping way: 
arrange(iris2, desc(Species), sepLength) %>% head

```

## 3. group_by() - equivalent to plyr::ddply()
```{r}
library(plyr)
# mutate iris2, replace iris2 with mutated data
iris2 %>% mutate(sepalRatio = sepWidth/sepLength, # a simple ratio
                 avgSepLen = mean(sepLength)) -> iris2 # can use mutate to do column-wise operations too

# transumute only outputs new columns
transmute(iris2, sepalRatio = sepWidth/sepLength)

# group the data by column "Species" - summarise by mean sepal and petal lengths
iris2 %>% group_by(Species) %>% summarise(mean(sepLength), mean(petLength))

# how you would do this in plyr; I think it makes more sense to use a single function call
# syntax: ddply(dataframe, groups, special function, function calls)
head(ddply(iris2, .(Species), summarize, 
      avgSepLen = mean(sepLength)))

# transform adds column(s) all at once
head(ddply(iris2, .(Species), transform, 
      avgSepLen = mean(sepLength)))

# mutate lets you adds column(s) sequentially, so they can be referenced internally within the function
head(ddply(iris2, .(Species), mutate, 
      avgSepLen = mean(sepLength), ratko = avgSepLen*10)) # ratko references the previous column: avgSepLen

# can we use a pipe to plot? Si se puede, a few examples below: 
# plot the groups on the same panel, using color to identify groups
iris2 %>% group_by(Species) %>% ggplot(aes(sepLength,petLength, color = Species)) + geom_point() + geom_smooth() 

# facet_wrap the different groups onto different panels
iris2 %>% group_by(Species) %>% ggplot(aes(sepLength,petLength, color = Species)) + geom_point() + geom_smooth() + facet_wrap(~Species)

# sfsmisc as an alternative for plotting?

# what about modelling; Can we fit a linear model to each group, then examine the summary data? - not sure how to use this yet...
iris2 %>% group_by(Species) %>% do(model = lm(sepLength ~ petLength, data = .)) -> iris.mods

# look at the structure - HOW DO I GET TO YOU?
str(iris.mods[1,2])

```
