---
title: "dplyr_session2"
author: "Cody Flagg"
date: "Monday, May 04, 2015"
output: html_document
---

```{r echo=FALSE,warning=FALSE,message=FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
```

# Outline

1. rename() : rename columns in a data frame
2. arrange() : order rows
2. group_by() : summary statistics by groups you define, like ddply(data, grp, transform)
3. Piping/Chaining %>% : connect manipulations into a single call
  - http://www.sharpsightlabs.com/dplyr-intro-data-manipulation-with-r/
  - https://rpubs.com/justmarkham/dplyr-tutorial
  - https://rpubs.com/mkapur/ModelingProject

## Future Topics
4. SQL functions
    - Connecting to databases (http://cran.r-project.org/web/packages/dplyr/vignettes/databases.html)
    - Manipulating data from DBs 
5. Two-table joins : multi-table data manipulation (http://cran.r-project.org/web/packages/dplyr/vignettes/two-table.html)
6. Data Cleaning (http://stackoverflow.com/questions/23642811/replace-parts-of-a-variable-using-numeric-indices-in-dplyr-do-i-need-to-create?rq=1)

## 1. rename() - rename columns

```{r}
# Let's play with the iris dataset for a bit, RA Fisher first analyzed these data in 1936
# first print the names of the columns
names(iris)
# The periods in each column name are kind of annoying, let's fix that
rename(iris, petLength = Petal.Length, 
  petWidth = Petal.Width, 
  sepWidth = Sepal.Width, 
  sepLength = Sepal.Length) -> iris2 # turn the assignment arround, remember this for piping later

# rename and select have special functions for finding columns
select(iris, contains("Sepal")) %>% head

# another example - notice that it is not case sensitive
select(iris, ends_with("width")) %>% head

# what if you have lots of columns, named/numbered something?
# create a silly example
df1 = data.frame(x1 = rnorm(5), x2 = runif(5), x3 = rpois(5, 2), type = sample(letters, 5))
# now select a few columns
select(df1, num_range("x", 1:2))
```


## 2. arrange() - re-order rows by column

```{r}
# arrange(dataframe, column1, column2, ...)
arrange(iris2, desc(Species), sepLength)
```

## 3. group_by() - equivalent to plyr::ddply()
```{r}
# iris2 %>% mutate(sepalRatio = sepWidth/sepLength) -> iris2

# group the data by column "Species" - summarise by mean sepal length 
group_by(iris2, Species) %>% summarise(mean(sepLength), mean(petLength))

group_by(iris2, Species) %>% ggplot(aes(sepLength,petLength)) + geom_point() + geom_smooth()



```
