---
title: "photoPaths_2015"
author: "Cody Flagg"
date: "Thursday, April 09, 2015"
output: pdf_document
---

# Summary
Craig Freeman at the USDA is putting together a "database" of photosynthetic pathways for plants in the USDA Plants list. There are currently 86,000 species listed in it, however the majority (>80%) have unknown pathways as of right now. Nicole Stanton, a field tech with D06, shared the list with me. There are weird symbols and inappropriate DB practices (i.e. violations of first normal form) that make the list less useful/query-able. I have cleaned it up with this script, and made it easier to search. 


## Setup: Import Data from XLS workbooks and Load Libraries
```{r setup}
# set working dir
setwd("C:/Users/cflagg/Documents/FSU_Projects_cflagg/FOPS Tools/photosynthetic_pathway_data_USDA")

# load libraries
library(XLConnect)
library(plyr)
library(dplyr)
library(stringr)
library(splitstackshape)

# load a list of files in the directory
fileGrab <- list.files(getwd(), full.names=TRUE) # list all the files, full.names=TRUE is necessary for ldplay/lapply to work below

# I added a prefix to the files to make it easier
prefix <- 'pp_'

# only grab files with the prefix
xlsGrab <- fileGrab[grep(prefix,fileGrab)] # subset to just the ones in your module, using prefix

```

## Read and Combine - Part 1
### Species-level pathway information 
```{r read1}
# read in .xlsx files
species.pp <- loadWorkbook(xlsGrab[2])
# read workbooks 
species.pp1 = readWorksheet(species.pp, sheet = 1, header = TRUE)
species.pp2 = readWorksheet(species.pp, sheet = 2, header = TRUE)
# combine into a single dataframe
species.pp.all = rbind.fill(species.pp1,species.pp2)
# rename columns
colnames(species.pp.all) <- c("taxonID", "scientificName","commonName","family","pathway","refType","reference")
# dump the last blank column
species.pp.all <- species.pp.all[,-8]
```

## Clean-up Data
* should copy scientificName to another column, then split column into genus and species (make it more query-able)
* also have to remove weird symbols from commonName (if commonName left 2 char = ">>", move string to new column "synonym")
* there are x's in scientificName that denote "certainty" for the pathway (certainty = a new column)
* forget about making spp./var. a 

### Execute full clean-up run
```{r clean1}
# add a column for synonyms
species.pp.all$synonym <- " " 

# Separate out symbols from commonName; if it >> symbols, put it in the synonym column
species.pp.all2 <- species.pp.all

species.pp.all2$synonym <- ifelse(str_sub(species.pp.all2$commonName, 1,2) == ">>", 
                        (species.pp.all2$synonym <- str_sub(species.pp.all2$commonName,3,nchar(species.pp.all2$commonName))) , 
                        (species.pp.all2$synonym = " "))

# make commonName blank if  there is a synonym (since this is not the actual common name)
species.pp.all2$commonName <- ifelse(species.pp.all2$synonym != " ", " ", species.pp.all2$commonName)

# mark the genusCertainPath column if there is an x in the scientificName
species.pp.all2$genusCertainPathway <- ifelse(str_sub(species.pp.all2$scientificName,1,1) == "x", "genus inferred", " ") 

###### THIS PART DROPS THE FIRST CHARACTER
# drop the x from the scientificName
species.pp.all2$scientificName <- ifelse(str_sub(species.pp.all2$scientificName,1,1) == "x" , 
                                         str_sub(species.pp.all2$scientificName,2,nchar(species.pp.all2$scientificName)), str_sub(species.pp.all2$scientificName,0,nchar(species.pp.all2$scientificName)))

species.pp.all2$scientificName <- str_trim(species.pp.all2$scientificName, "left")

# split the names up by spaces
species.pp.all3 = cSplit(species.pp.all2, "scientificName", sep = " ", fixed = TRUE)
# cSplit returns a data.table, which changes how you can index rows and columns...
species.pp.all3 <- as.data.frame(species.pp.all3)

# preserve this column
species.pp.all3$scientificName <- species.pp.all2$scientificName

# cSplit turns the new vectors into factors - change to characters
species.pp.all3[] <- lapply(species.pp.all3, as.character)

# replace NA's with blanks, else the concatentated name includes NA
species.pp.all3[is.na(species.pp.all3)] <- ""

# dump extra columns
http://stackoverflow.com/questions/5992082/how-to-remove-all-whitespace-from-a-string
species.pp.all4 <- subset(species.pp.all3, select = c(1:10,32))

# trim the trailing whitespace on the right-side, uses stringr
species.pp.all4$synonym<-str_trim(species.pp.all3$synonym, "left")

# rename specific columns
setnames(species.pp.all4, c(9:10), c("genus","species"))

# write to a new file
write.csv(species.pp.all4, file = "photo_pathways_clean.csv")
```


## Only grab graminoids

```{r}
# read in the file from the wd
gram <- read.csv(fileGrab[2], header=T)

# Check the spelling
summary(gram$family)

# using %in% rather than == will leave out NAs
gram.only <- gram[gram$family %in% c("Poaceae", "Cyperaceae", "Juncaceae"),]

write.csv(gram.only, file = "photo_pathways_graminoidsOnly.csv")
```
