---
title: "Loops and Apply"
author: "E Pluribus Unum"
date: "March 8, 2018"
output: html_document
---

# Outline

* Intro to loops
* 1. Types of loops
* 2. Common Application
* 3. Nested loops
* 3. Apply solutions
* 4. Kristina's example

```{r}
### What's a loop?

# Synonyms: cycle, iteration, replicants

#"It is nothing more than automating a multi-step process by organizing sequences of actions or 'batch' processes and by 
# grouping the parts that need to be repeated." - some website

# Loops allow you to repeat stuff!

### Three control flow commands: while, repeat, and for
# "Control commands enable a project to branch between alternatives, or to "take descision", so to speak" - some website

# While and repeat are verification-based commands. They follow rules like, "keep going until a condition is met"
# the for command is a counter, or index-based command. It keeps going until it runs out of things. Here are some examples:

### Two clauses: break and next
 # break tells the loop to stop when conditions are met
 # next skips to the next iteration
```

## PART 1: Types of loops

```{r}
### FOR LOOP
for(ii in 1:10){
  print(ii)
}
```

```{r}
### WHILE LOOP
ii <- 0
while(ii != 10){
  print(ii)
  ii <- ii + 1
}
```

```{r}
### REPEAT LOOP
ii <- 0 
repeat{
  print(ii)
  ii <- ii + 1
  if(ii == 10) break
}
```


* How would you verbally descibe what each of these loops is doing? 


* TRY IT YOURSELF!
* Make a for-loop that counts backwards from 100, but stops at 10

## PART 2: common application

* Imagine you have 50 plots with about 20 plant height measurements per plot, and you want to find the per plot mean height
* The below for-loop will cycle through each plot and find the mean for that plot's measurements. Then it'll put the 
* measurements together in a data.frame

```{r, eval=FALSE}
dat <- data.frame(plotID = 1:50, height = rnorm(1000, 10, 10)) # Generate dummy dataset

## best to avoid assigning an object name to a global function... i.e. "plot()" is the default graphing function
plots <- unique(dat$plotID)                            # first we need a list of each plot, with no repeats. 

for(ii in 1:length(plots)){                       # tell the for-loop to run the same number of times as the number of plots we have
  a <- subset(dat, dat$plotID == ii)               # subset the data.frame "dat" to include only rows where plot matched "ii"
  b <- mean(a$height)                            # find the mean height for plot "ii"
  c <- data.frame(plotID = dat$plotID[ii], mean = b) # put the mean height and plot id into a data.frame
  if(ii == 1){                                   # we now need to compile each row of results (id and mean height) into a data.frame. To do this, we first check if this is the first time the for-loop has run by testing ii == 1
    d <- c                                       # if this is the first run, then make "d" equal to "c"
  }else{
    d <- rbind(d, c)                             # if false, then make "d" equal to "c" and "d" combined
  }
}
print(d)
```


### Try it yourself!
* imagine you have 8000 kittens, and you need to find out the proportion of kittens that have missing tails by size class.
* you should end up with a 2 column data.frame, 5 rows long


```{r}
dat <- data.frame(sizeclass = sample(1:5, 8000, replace=TRUE), tail = sample(c("yes", "no"), 8000, replace=TRUE)) # Generates dummy data
```

########## PART 3: Nested Loops ########## 

```{r}
# Generate dummy data
dat <- data.frame(site = sample(1:5, 1000, replace=TRUE), year = sample(2001:2010, 1000, replace=TRUE), contamination = rnorm(1000, 0.1, 1))

# Imagine you're looking at water contamination at five sites over the past ten years. You want to find the mean  
# contamination levels per site, per year. Sounds like the job for a nested loop!

sites <- unique(dat$site)                                                 # make list of sites and years
years <- unique(dat$year)
for(ii in 1:length(sites)){
  a <- subset(dat, dat$site == sites[ii])                                # subset to one site
for(jj in 1:length(years)){                                              # note that we use a different variable, "jj" for the nested loop
  b <- subset(a, a$year == years[jj])                                    # subset again to one year for that one site
  c <- mean(na.omit(b$contamination))                                    # find mean contamination
  d <- data.frame(site = sites[ii], year = years[jj], contamination = c) # put data into a data.frame
  if(jj == 1){
    e <- d
  }else{
    e <- rbind(d, e)
  }
}
  if(ii == 1){
    f <- e
  }else{
    f <- rbind(f, e)
  }
  }
results <- f
print(head(results, 5))
```

## Loop optimization and debugging advice

* If you're just trying to summarize data, just use `apply()`, `aggregate()`, `plyr`, or `dplyr` functions
    * People have figured out how to avoid complex loops for summarizing data in R, use their solutions
    * Loops are well suited for conditional logic problems that require `ifelse()` type of cases
* Use a print statement for every "transformation" or nested level
* Learn to use the `browser()` and `debug()` functions if something isn't working
* Fix your expectations
    * Have a concrete idea of what the first line and last line of output should be, then check them
    * If they aren't what you expected, inspect the loop
* If you know the  dimensions of your output object, [pre-allocate](https://stackoverflow.com/questions/11486369/growing-a-data-frame-in-a-memory-efficient-manner) it before running the loop
    * This will decrease the run time
* Global, "naked" for loops are poor practice
    * Enclose a for loop within a function to debug and re-use
    * Dumping objects into your global environment can be confusing and affect other snippets of code
    
```{r}
## a non-global conditional logic for loop run with browser() on something that isn't a data.frame/data table

## practice data set, an ugly non-tabular thing with mixed data types (numbers, characters, data frame, list etc.)
## This is an example of messy data you might find from the internet...or from a collaborator
## a named list, where the names are a, b, c, d, e
test <-  list(a = c(1:10), b = c(LETTERS),c = list(LETTERS), d = list(c(1:10, rnorm(100))), e = data.frame(x = rep(10), y = rnorm(10)))

## problem: how many items do I have in each element of this list, what is the data type, and are there any lists or data frames nested in it?

my_loop = function(){
  #browser() # uncomment this and run to see what browser() does
  output_object = data.frame(x = 10, y = 10) ## pre-allocated object
  for (things in names(test)){
    
    print(things)
    pointer = test[things][[1]] ## accessing list element
    
    if(is.character(pointer)){
      print("it's not a number")
    } else if (is.numeric(pointer)) {
      print("itza a number!") 
    } else if (is.data.frame(pointer)){
      ## You want the data table, "export" it for further use
      print("found a data frame!")
      output_object = pointer
      return(output_object)
    } else {
      # everything else...
      print("...must be a list")
    }
  }
} 

## How is this "non global"? 
# the object "pointer" was assigned at each iteration of the for loop, type it here to see what it is
# pointer

## run your non global loop any time in one line
my_loop()

## assign the output of your functional loop
outputz <- my_loop() 
```


## apply and apply-like solution alternatives (from Cody)

* `apply` functions are an alternative to for loops
* `apply` functions actually use for-loops "under the hood" i.e. you'll see for loops if you look at the source code
    * Base R apply functions are [often written in C](http://adv-r.had.co.nz/C-interface.html); writing code in C, then calling it from R, happens in a lot of functions and can be faster in a lot of cases
* __Old Perception__: ["for loops in R are slow and should be avoided"](https://stackoverflow.com/questions/7142767/why-are-loops-slow-in-r)
   * Check Hadley Wickham's [section on "Loops" (ctrl+f "loops")](http://adv-r.had.co.nz/memory.html) for an explanation of why for loops _can_ be slow
* __Today__: for loops can be just as ["performant"](https://datascienceplus.com/strategies-to-speedup-r-code/) or fast in R now, and were optimized many years ago
    * There are a few useful best practices to follow for writing optimal for loops:
        * pre-allocate memory e.g. if your output `data.frame` will be 50 rows and 3 columns...create an empty `data.frame` with these dimensions before running the for-loop
        
### Why use `apply` and `apply-like` functions? 

* Readability: write code in a linear, "top to bottom" style (whereas reading a for loop requires "jumping" up and down nested levels)
* Shorter code
* Don't have to track iterator tokens and indices for highly nested problems (i.e. `for (x in vector) {for (y in subvector) { for (z in subsubvector){mydata[x, y, z]}}}`)
* Fewer side effects

### Solutions 

* `plyr` and `dplyr` are modern, multi-variate solutions
    * Written by an extremely thoughtful person who is now the [Chief Scientist at RStudio](http://hadley.nz/) ...among other things
    * [Part of the "Hadley-verse" or "tidyverse": a collection of related, interoperable R packages](https://www.tidyverse.org/)
* `plyr` has similar functionality to Base R apply functions but more...
    * Support for working with all major R data structures (vector, list, data.frame, matrix)
    * Cleaner syntax (function arguments have the same names)
    * Simple function grammar 
        * ("ddply"" = data frame input, data frame output; "ldply"" = list input, data frame output; "dlply"" = data frame input, list output etc.)
    * Easier multivariate/highly nested processing 
        * e.g. how fun will it be to write a nested for loop to summarize a data set with 5 factor columns and one response variable?
            * plyr: `ddply(my_data, ~column1+column2+column3+column4+column5, summarize, mean_response = mean(my_response_variable))`
            * for loop: `for u..., for v..., for y..., for x..., for z... {kill_me_now[i]}`
    * Check this [link](https://github.com/cflagg/codingSupportGroup/blob/1574132dcfaff373e199598ab09acf985173f9a5/Lessons/dataWrangling/plyr_session/FSU_plyr_dplyr_session_04172015.Rmd) for a table of how apply functions map to plyr functions (very straightforward)


#### Cat Tails Problem

```{r}
library(plyr)

## same data as above
katz <- data.frame(sizeclass = sample(1:5, 8000, replace=TRUE), tail = sample(c("yes", "no"), 8000, replace=TRUE)) # Generates dummy data

## a plyr approach
## plyr expands the apply function universe into multi variate problems

## "dd = data frame to data frame"
## Plain English: "in data frame 'katz', for each unique level/factor in sizeclass (~sizeclass), do the following and return a data frame...
ddply(katz, ~sizeclass, summarize, 
      ## count number of rows where tail == no
      no_tail = length(which(tail == "no")),
      ## count number of rows where tail == yes
      tail = length(which(tail == "yes")),
      ## calculate proportion for each level of sizeclass without tail""
      proportion_no_tail = no_tail/(tail+no_tail))
      ## give me a new data frame with columns [sizeclass, no_tail, tail, proportion_no_tail]
```

* `dplyr` is the "successor" package to plyr, but is geared towards only working with data.frames/tabular data 
    * Common syntax
    * Function grammar borrows from SQL query language
    * String multiple data transformations together without the intermediate data objects

```{r}
library(dplyr)
library(tidyr)
## dplyr approach
## need tidyr to go from long to wide structure
katz %>% group_by(sizeclass, tail) %>% summarize(countz = n()) %>% spread(tail, countz) -> katz2
katz2 %>% summarize(no_tail = no/(no+yes))
```

####apply solution to contamination mean problem

```{r}
tapply(dat$contamination, list(dat$site, dat$year), mean)
```

### plyr solution to contamination problem

```{r}
## mixing plyr (ddply) with dplyr (arrange) functions
ddply(dat, ~site+year, summarize, mean_contam = mean(contamination)) %>% arrange(year)
```


## PART 4: Kristina's Data

* Note: I didn't provide the files to execute this - I'll show you how it works in-person

```{r, eval=FALSE}
setwd("LEFT BLANK                     ")       # folder that includes all of the files, and only them
files <- list.files()                          # makes list of file names in that folder
for(ii in 1:length(files)){                    # the loop!
  a <- files[ii]
  b <- read.csv(a)                             # we load each file into R one at a time
  Multitavelength <- substr(b[2,1], 28, 40)    # these isolate the text we want on each line/row
  Test_name <- substr(b[3,1], 31, 40)
  Date_Cell_Correction <- substr(b[4,1], 28, 40)
  Measurement_mode <- substr(b[5,1], 31, 40)
  Sample_positioner <- substr(b[6,1], 35, 40)
  Number_of_Samples <- substr(b[7,1], 40, 40)
  Cell_correction <- substr(b[8,1], 39, 40)
  Setup_correction <- substr(b[9,1], 39, 40)
  ID <- substr(b[10,1], 40, 40)
  Auto_save_data <- substr(b[11,1], 39, 40)
  Data_file_name <- substr(b[12,1], 38, 40)
  nm488 <- as.numeric(substr(b[17,1], 14, 40))
  nm1000 <- as.numeric(substr(b[18,1], 14, 40)) # adding "as.numeric" here produces "NA" when no measuremnet was taken
  c <- data.frame(Multitavelength = Multitavelength, Test_name = Test_name, Date_Cell_Correction = Date_Cell_Correction, 
                  Measurement_mode = Measurement_mode, Sample_positioner = Sample_positioner, Number_of_Samples = Number_of_Samples,
                  Cell_correction = Cell_correction, Setup_correction = Setup_correction, ID = ID, Auto_save_data = Auto_save_data,
                  Data_file_name = Data_file_name, nm488 = nm488, nm100 = nm1000) # this puts the data into a singe row data.frame
  if(ii == 1){                                  # this puts the data.frame rows together
    d <- c
  }else{
    d <- rbind(d, c)
  }
}
results <- d
print(results) # yay!

# save.csv(results, "results.csv") # this will save the results as a csv in the same folder. 

#



########## PART 5: Functions ###########
### Functions are like little programs in your r script - they help automate processes

fun <- function(x){
  a <- x * 2
  print(a)
}
fun(10)


### Example function: standard error of the mean
sem <- function(x){
  sd(x)/sqrt(length(x))
}

sem(c(10, 5, 1))



### Example function: Size-mass index
smi <- function(m, l){
  require(lmodel2)
  b <- lmodel2(m ~ l)$regression.results[3,3]
  print(m*(mean(na.omit(l))/l)^b)
}

mass   <- sample(1:100, 100)
length <- sample(1:100, 100)

smi(mass, length)


### Example function: Comparative modeling table
getstr = function(mystring, initial.character, final.character){
  if (!is.character(mystring)) { stop('The parent string must be a character variable.') }
  if (!is.character(initial.character)){ stop('The initial character must be a character variable.') }
  if (!is.character(final.character)) { stop('The final character must be a character variable.') }
  snippet = rep(0, length(mystring))
  for (i in 1:length(mystring)){ initial.position = gregexpr(initial.character, mystring[i])[[1]][1] + 1
  final.position = gregexpr(final.character, mystring[i])[[1]][1] - 1
  snippet[i] = substr(mystring[i], initial.position, final.position)}
  return(snippet)
}


AICtable <- function(mods){
  require(MuMIn)
  require
  for(ii in 1:length(mods)){
    AICcs <- AICc(mods[[ii]])
    k <- length(summary(mods[[ii]])$coefficients[,1])-1
    names <- getstr(as.character(summary(mods[[ii]])[1]), "~", ",")
    x <- data.frame(names, AICcs, k)
    if(ii == 1){
      xx <- x
    }else{
      xx <- rbind(xx, x)
    }
  }
  xx$deltaAICc <- NA
  xx$deltaAICc <- round(xx$AICc - xx$AICc[1], 3)
  xx$deltaAICc <- xx$deltaAICc + abs(min(xx$deltaAICc))
  xx$wi <- exp(-0.5*xx$deltaAICc) # note that the package Mulm includes better methods for estimating AICc and wi (don't use my bad math)
  xx$wi <- round(xx$wi/sum(xx$wi), 3)
  colnames(xx) <- c("model", "AICc", "k", "deltaAICc", "wi")
  xx
}

dat <- data.frame(a = rnorm(10, 10, 10), b = rnorm(10, 5, 5), c=rnorm(10, 20, 20), d=rnorm(10, 1, 2)) # generate dummy data

mod1 <- glm(a ~ b * c * d, data = dat)
mod2 <- glm(a ~ b * c + d, data = dat)
mod3 <- glm(a ~ b + c * d, data = dat)
mod4 <- glm(a ~ b + c + d, data = dat)
mod5 <- glm(a ~ b + c    , data = dat)
mod6 <- glm(a ~ b +     d, data = dat)
mod7 <- glm(a ~     c + d, data = dat)
mod8 <- glm(a ~ b +      , data = dat)
mod9 <- glm(a ~     c    , data = dat)
mod10 <- glm(a ~        d, data = dat)
mod11 <- glm(a ~ 1       , data = dat)

mods <- list(mod1, mod2, mod3, mod4, mod5, mod6, mod7, mod8, mod9, mod10, mod11)

AICtable(mods)

#
```